---
title: "Homework_11"
date: "4/8/2020"
author: "Cai McCann"
output: 
  html_document:
      toc: true
editor_options: 
  chunk_output_type: console
---

--------------------------------------------------------------------------------

# Batch Processing Question: 

Repeat the exercise from the Batch Processing Lecture (7th April), but do it using real data sets rather than purely simulated. Check with folks in your lab to see if there are multiple data sets available for analysis. Stick to simple data analyses and graphics, but try to set it up as a batch process that will work on multiple files and save summary results to a common file.

### Function for linear regression analysis and summary results

```{r linear regression function} 

# ----------------------------------------------------------
# FUNCTION reg_stats
# description: fits linear models, extra model stats
# inputs: 2-column data frame (x and y)
# outputs: slope, p-value, and r2
############################################################
reg_stats <- function(d=NULL) {
            if(is.null(d)) {
              cbmi <- runif(150) 
              lsystolic <- runif(150) 
              d <- data.frame(cbmi,lsystolic)
            }  
  . <- lm(data=d,d[,2]~d[,1])
  . <- summary(.)
  stats_list <- list(Slope=.$coefficients[2,1],
                     pVal=.$coefficients[2,4],
                     r2=.$r.squared)
  
return(stats_list)

} # end of reg_stats
# ----------------------------------------------------------

```


### Set up and Batch Process in Data

```{r setup and batch process, warning=FALSE, message=FALSE}
# Preliminaries ---------------------------------------------------------
# load packages
library(ggplot2)
library(TeachingDemos)

char2seed("spring snow")

# Global variables ---------------------------------------------------------
file_folder <- "bp_bmi_analysis/" 
n_files <- 3
file_out <- "BMIbpStatsSummary.csv"
file_names <- list.files(path=file_folder)

# Create a data frame to hold summary file statistics
ID <- seq_along(file_names)
f_name <- file_names
slope <- rep(NA,length(f_name))
p_val <- rep(NA,length(f_name))
r2 <- rep(NA,length(f_name))

stats_out <- data.frame(ID,f_name,slope,p_val,r2)

#########################################################################
# initiate batch process to loop through individual files

for (i in seq_along(file_names)) { # starts for loop in the function
  data <- read.table(file=paste(file_folder,file_names[i],
                                sep=""),
                                sep=",",
                     header=TRUE)
  # apply calculation and extract results
  . <- reg_stats(data) # pull out regression stats from clean file
  stats_out[i,3:5] <- unlist(.) # unlist, copy into last 3 columns
} # end of for loop


# set up data file and incorporate time stamp and minimal metadata
write.table(cat("# Summary statistics for",
                " linear regression batch processing", 
                "\n",
                "# timestamp: ",
                as.character(Sys.time()),
                "\n",
                "# CLM", 
                "\n",
                "# ____________________", 
                "\n",
                "\n",
              file=file_out,
              row.names="",
              col.names="",
              sep=""))
                
# now add the data frame
write.table(x=stats_out,
            file=file_out,
            sep=",",
            row.names=FALSE,
            append=TRUE)

# ----------------------------------------------------------

print(stats_out)

```

*Here is the resulting output CSV for reference:*

```{r, echo=FALSE}
# installed package to be able to embed this: 
# xfun::pkg_load2(c("htmltools", "mime"))

# a single file of the output statistics
xfun::embed_file('BMIbpStatsSummary.csv')

```