---
title: "Homework_5"
author: "Cai McCann"
date: "2/12/2020"
output: 
  html_document:
      toc: true
---

--------------------------------------------------------------------------------

### Question 1:

Assign to the variable n_dims a single random integer between 3 and 10.

-	Create a vector of consecutive integers from 1 to n_dims2n_dims2.

-	Use the sample function to randomly reshuffle these values.

-	create a square matrix with these elements.

-	print out the matrix.

-	find a function in r to transpose the matrix.

-	print it out again and note how it has changed.

-	calculate the sum and the mean of the elements in the first row and the last row.

-	read about the eigen() function and use it on your matrix

-	look carefully at the elements of values and vectors. What kind of numbers are these?

-	dig in with the typeof() function to figure out their type.

-	if have set your code up properly, you should be able to re-run it and create a matrix of different size because n_dims will change.


```{r q1}
# assign to the variable n_dims a single random integer between 3 and 10.
n_dims <- sample(3:10,1) #sample(range,# of integers)
print(n_dims)

# create a vector of consecutive integers
# from 1 to n_dims^2.
x <- seq(from=1,to=n_dims^2)
print(x)

# use the sample function to randomly reshuffle these values
x_reshuffled <- sample(x)
print(x_reshuffled)
# create a square matrix with these elements
m <- matrix(nrow=n_dims, ncol=n_dims, data=x_reshuffled) 
  #set rows and cols to original variable for repeatibility

# print out the matrix.
print(m)

# find a function in r to transpose the matrix
m2 <- t(m)

# print it out again and note how it has changed.
print(m2)
  # verified the change of rows and columns in this matrix

# calculate the sum and the mean of the elements in the first row and the last row

mean_firstrow <- mean(m2[1,]) #selects elements in column 1 of variable and averages 
sum_firstrow <- sum(m2[1,])
print(mean_firstrow)
print(sum_firstrow)

mean_lastrow <- mean(m2[n_dims,]) #set n_dims number of rows for repeatibility
sum_lastrow <- sum(m2[n_dims,])
print(mean_lastrow)
print(sum_lastrow)

# read about the eigen() function and use it on your matrix; typed: ?eigen
# typeof(m2) #the usual matrix elements are integers

m_eigen <- eigen(m2) # eigen transformed numbers
# look carefully at the elements of $values and $vectors. What kind of numbers are these?
# dig in with the typeof() function to figure out their type
typeof(m_eigen$values) 
typeof(m_eigen$vectors) 
  # $values and $vectors elements are complex (imaginary numbers)

```


### Question 2:

Create a list with the following named elements:

- my_matrix, which is a 4 x 4 matrix filled with random uniform values

- my_logical which is a 100-element vector of TRUE or FALSE values. Do this efficiently by setting up a vector of random values and then applying an inequality to it.

- my_letters, which is a 26-element vector of all the lower-case letters in random order.

- create a new list, which has the element[2,2] from the matrix, the second element of the logical vector, and the second element of the letters vector.

- use the typeof() function to confirm the underlying data types of each component in this list

- combine the underlying elements from the new list into a single atomic vector with the c() function.

- what is the data type of this vector?


```{r q2}
# create a new list, which has the element[2,2] from the matrix, the second element of the logical vector, and the second element of the letters vector.
my_list <- list(my_matrix = matrix(runif(16), ncol = 4, nrow = 4),
                my_logical = runif(100)>0.5,
                my_letters = sample(letters[1:26]))
                
  # my_matrix generates 16 random values to fix a matrix of 4 columns and 4 rows 
  
  # my_logical is a 100-element vector of TRUE or FALSE values--stores x variable's 100 values that are greater than 0.5, thus creates BOOLEAN  
  
  # my_letters, which is a 26-element vector of all the lower-case letters in random order.
  # randomly samples 26x from letters function
print(my_list)

# create a new list, which has the element[2,2] from the matrix, the second element of the logical vector, and the second element of the letters vector.
new_list <- list(my_list$my_matrix[2,2], my_list$my_logical[2], my_list$my_letters[2])
print(new_list)

# use the typeof() function to confirm the underlying data types of each component in this list
#*queries data type by selecting list element from another list 
typeof(new_list[[1]]) # double
typeof(new_list[[2]]) # logical
typeof(new_list[[3]]) # character

# combine the underlying elements from the new list into a single atomic vector with the c() function.
#*stores concatenated list element from another list 
combined_list_vec <- c(new_list[[1]],new_list[[2]],new_list[[3]])
print(combined_list_vec)

# what is the data type of this vector?
typeof(combined_list_vec)#character
# per hierarchy of coercion, character is lowest and is what typeof queried

```


### Question 3:

- Create a data frame with two variables (= columns) and 26 cases (= rows).
call the first variable my_unis and fill it with a random uniform value from 0 to 10

- call the second variable my_letters and fill it with 26 capital letters in random order.

- for the first variable, use a single line of code in R to select 4 random rows and replace the numerical values in those rows with NA.

- for the first variable, write a single line of R code to identify which rows have the missing values.

- for the second variable, sort it in alphabetical order

- calculate the column mean for the first variable.

```{r q3}
## Create a data frame with two variables (= columns) and 26 cases (= rows).

# call the first variable my_unis and fill it with a random uniform value from 0 to 10
my_unis <- runif(26,min=0,max=10) #obtain 26 random numbers
print(my_unis)

# call the second variable my_letters and fill it with 26 capital letters in random order.
my_letters <- sample(LETTERS[1:26]) #samples randomly CAPITALIZED letters
print(my_letters)

# created that data frame, homestretch!
d_frame <- data.frame(my_unis,my_letters) #add variables to dataframe
print(d_frame) #view all data
head(d_frame) #peek at data
str(d_frame) #view metadata, e.g. nrow

# for the first variable, use a single line of code in R to select 4 random rows and replace the numerical values in those rows with NA.
d_frame[(sample(nrow(d_frame), size = 4, replace = FALSE)), 1] <- NA
  #assign NA to random sample of 4 row values, w/o replacement
print(d_frame)

# for the first variable, write a single line of R code to identify which rows have the missing values.
which(is.na.data.frame(d_frame)==TRUE)
  #query and print which rows that NA is TRUE=present
 
# for the second variable, sort it in alphabetical order
sort(d_frame$my_letters)

# calculate the column mean for the first variable.
mean(d_frame[which(is.na.data.frame(d_frame[,1])==FALSE),1])
    #average from selected subset which values are NOT "NA", by row, w/o replacement
# this is another helpful option: mean(d_frame$my_unis, na.rm=TRUE)

```

