---
title: "Homework_6"
author: "Cai McCann"
date: "2/19/2020"
output: 
  html_document:
      toc: true
  chunk_output_type: console
---

--------------------------------------------------------------------------------

```{r set up, warning=FALSE}
#load library
library(ggplot2) # for graphics 
library(MASS) # for maximum likelihood estimation
```

### Simulated Data

``` {r fake data}
# generate some fake data here. Quick and dirty, a truncated normal distribution to work on the solution set
z <- rnorm(n=3000,mean=0.2) #create a random normal distribution with 3000 observations and a mean of 0.2 and assign it to z variable
z <- data.frame(1:3000,z) #convert number set 1 to 3000, z to be used for columns 
names(z) <- list("ID","myVar") #assign row and column names "ID" and "myVar" respectively to the z vector
z <- z[z$myVar>0,] #from the z vector, select the only the myVar rows >0 and subset it to z
str(z) #Compactly Display the Structure of z
summary(z$myVar)

```

### Plotting Simulated Data

```{r plots of fake data, message=FALSE, warning=FALSE}
# Plot histogram of data (preset to y = ..density.. for area under curve of 1)
p1 <- ggplot(data=z, aes(x=myVar, y=..density..)) +
  geom_histogram(color="grey60",fill="cornsilk",size=0.2) 
print(p1)

# Add empirical density curve (creates smoothed, fitted distribution)
p1 <-  p1 +  geom_density(linetype="dotted",size=0.75)
print(p1)

# Get maximum likelihood parameters for normal function 
normPars <- fitdistr(z$myVar,"normal")
print(normPars)
str(normPars)
normPars$estimate["mean"] # note structure of getting a named attribute

# Plot normal probability density--symmetrical and poor fit due skewed data
meanML <- normPars$estimate["mean"]
sdML <- normPars$estimate["sd"]

xval <- seq(0,max(z$myVar),len=length(z$myVar))

stat <- stat_function(aes(x = xval, y = ..y..), fun = dnorm, colour="red", n = length(z$myVar), args = list(mean = meanML, sd = sdML))
p1 + stat

 # Plot exponential probability density--unusually high and poor fit as well
expoPars <- fitdistr(z$myVar,"exponential")
rateML <- expoPars$estimate["rate"]

stat2 <- stat_function(aes(x = xval, y = ..y..), fun = dexp, colour="blue", n = length(z$myVar), args = list(rate=rateML))
p1 + stat + stat2

# Plot uniform probability density--by its nature, a poor fit of these data
stat3 <- stat_function(aes(x = xval, y = ..y..), fun = dunif, colour="darkgreen", n = length(z$myVar), args = list(min=min(z$myVar), max=max(z$myVar)))
p1 + stat + stat2 + stat3
 
# Plot gamma probability density--relatively appropriate fit to the data
gammaPars <- fitdistr(z$myVar,"gamma")
shapeML <- gammaPars$estimate["shape"]
rateML <- gammaPars$estimate["rate"]

stat4 <- stat_function(aes(x = xval, y = ..y..), fun = dgamma, colour="brown", n = length(z$myVar), args = list(shape=shapeML, rate=rateML))
p1 + stat + stat2 + stat3 + stat4

# Plot beta probability density--adjusted to 0-1 and perhaps an appropriate fit
pSpecial <- ggplot(data=z, aes(x=myVar/(max(myVar + 0.1)), y=..density..)) +
  geom_histogram(color="grey60",fill="cornsilk",size=0.2) + 
  xlim(c(0,1)) +
  geom_density(size=0.75,linetype="dotted")

betaPars <- fitdistr(x=z$myVar/max(z$myVar + 0.1),start=list(shape1=1,shape2=2),"beta")
shape1ML <- betaPars$estimate["shape1"]
shape2ML <- betaPars$estimate["shape2"]

statSpecial <- stat_function(aes(x = xval, y = ..y..), fun = dbeta, colour="orchid", n = length(z$myVar), args = list(shape1=shape1ML,shape2=shape2ML))
pSpecial + statSpecial

```

### Clinical Dataset Comparison

This dataset comprises 2 columns and 1,238 rows of patient ids and corresponding BMI measurements. 

```{r my data}
# Read in my file
z <- read.table("C:\\Users\\clmcs\\Desktop\\BIOL381_McCann1\\clinical_cbmi_only.csv",header=TRUE,sep=",", stringsAsFactors=FALSE)
str(z)
summary(z)
head(z)
```

### Plotting Clinical Dataset

```{r plotting my data, warning=FALSE}

# Plot histogram of data--these data are slightly more symmetrical than previous 
p1 <- ggplot(data=z, aes(x=cbmi, y=..density..)) +
  geom_histogram(bins=50,
                 color="lightsteelblue",
                 fill="paleturquoise",
                 size=0.2) 
print(p1)

# Add empirical density curve
p1 <-  p1 +  geom_density(linetype="dotted",size=0.75)
print(p1)

# Get maximum likelihood parameters for normal function
normPars <- fitdistr(z$cbmi,"normal")
print(normPars)
str(normPars)
normPars$estimate["mean"] # note structure of getting a named attribute

# Plot normal probability density--not ideal fit to these slightly skewed data
meanML <- normPars$estimate["mean"]
sdML <- normPars$estimate["sd"]

xval <- seq(0,max(z$cbmi),len=length(z$cbmi))

stat <- stat_function(aes(x = xval, y = ..y..), fun = dnorm, colour="red", n = length(z$cbmi), args = list(mean = meanML, sd = sdML))
p1 + stat

# Plot exponential probability density--poor fit 
expoPars <- fitdistr(z$cbmi,"exponential")
rateML <- expoPars$estimate["rate"]

stat2 <- stat_function(aes(x = xval, y = ..y..), fun = dexp, colour="blue", n = length(z$cbmi), args = list(rate=rateML))
p1 + stat + stat2

# Plot uniform probability density--poor fit
stat3 <- stat_function(aes(x = xval, y = ..y..), fun = dunif, colour="darkgreen", n = length(z$cbmi), args = list(min=min(z$cbmi), max=max(z$cbmi)))
p1 + stat + stat2 + stat3

# Plot gamma probability density--gamma is slightly better than normal
gammaPars <- fitdistr(z$cbmi,"gamma")
shapeML <- gammaPars$estimate["shape"]
rateML <- gammaPars$estimate["rate"]

stat4 <- stat_function(aes(x = xval, y = ..y..), fun = dgamma, colour="brown", n = length(z$cbmi), args = list(shape=shapeML, rate=rateML))
p1 + stat + stat2 + stat3 + stat4

# Plot beta probability density--also somewhat better compared to normal
pSpecial <- ggplot(data=z, aes(x=cbmi/(max(cbmi + 0.1)), y=..density..)) +
  geom_histogram(color="grey60",fill="cornsilk",size=0.2) + 
  xlim(c(0,1)) +
  geom_density(size=0.75,linetype="dotted")

betaPars <- fitdistr(x=z$cbmi/max(z$cbmi + 0.1),start=list(shape1=1,shape2=2),"beta")
shape1ML <- betaPars$estimate["shape1"]
shape2ML <- betaPars$estimate["shape2"]

statSpecial <- stat_function(aes(x = xval, y = ..y..), fun = dbeta, colour="orchid", n = length(z$cbmi ), args = list(shape1=shape1ML,shape2=shape2ML)) 
pSpecial + statSpecial

# Find ML parameters given that Gamma probability density  is best (has to be positive)
gammaPars

```

### Best-fitting distribution and Simulation with Associated ML parameters

While it may appear gamma or beta distributions are options for these data, gamma does do well in that it is more realistic and does not have an upper bound the way beta does (largest data point = true upper bound). Therefore, I will choose the gamma distribution as my "best-fitting" distribution. 

```{r best fit gamma with simulation, warning=FALSE}

# Note: calculated ML parameters for Gamma probability density
# > gammaPars
#       shape         rate    
#   20.93801915    0.78475512 
#  ( 0.83526254) ( 0.03168299)

# Set up gamma probability density parameters and simulation
gammaPars <- fitdistr(z$cbmi,"gamma")
shapeML <- gammaPars$estimate["shape"]
rateML <- gammaPars$estimate["rate"]

z2 <- rgamma(n=nrow(z),shape=20.93801915,rate=0.78475512) # input specific parameters

d_z2 <- as.data.frame(z2) # make dataframe for graphing

str(z2) # confirmed vector length matches original

# Plot histogram of data
p_z2 <- ggplot(data=d_z2, aes(x=z2, y=..density..)) +
  geom_histogram(bins=50,
                 color="grey60",
                 fill="cornsilk",
                 size=0.2) 

# Plot Gamma probability density
stat4_z2 <- stat_function(aes(x = z2, y = ..y..), fun = dgamma, colour="blue", n = length(z2), args = list(shape=shapeML, rate=rateML))

p_z2 + stat4_z2 # layer gamma on histogram

# My original data's histogram and gamma distribution for comparison
p1 + stat4 

```

### Note on Comparison

The original and simulated histogram profiles appear relatively similar. The main difference is that the original data are narrower while the simulated data have a broader spread and are slightly more symmetrical. That said, I think the model is doing a good job of simulating realistic data that match the original measurements overall. 


