---
title: "Homework_8"
author: "Cai McCann"
date: "3/4/2020"
output: 
  html_document:
      toc: true
editor_options: 
  chunk_output_type: console
---

--------------------------------------------------------------------------------

### Question 1: Examine the structure of the iris dataset. How many observations and variables are in the dataset?

```{r q1, message=FALSE, warning=FALSE}
#make sure packaged loaded (can type in console as install.packages("tidyverse")
library(dplyr) #error, remove.packages("rlang"), reinstall

# peek at dataset called iris
glimpse(iris)
head(iris)

```

There are 150 observations and 5 variables in the dataset

### Question 2: Create a new data frame iris1 that contains only the species virginica and versicolor with sepal lengths longer than 6 cm and sepal widths longer than 2.5 cm. How many observations and variables are in the dataset?

```{r q2}
# subset iris dataset for specific parameters of Species, Sepal.Length, and Sepal.Width
iris1 <-filter(iris, Species %in% c("virginica","versicolor"), Sepal.Length > 6, Sepal.Width > 2.5)

# peek again
glimpse(iris1)
head(iris1)
summary(iris1$Species)

```

There are 56 observations and 5 variables in the dataset.

### Question 3: Now, create a iris2 data frame from iris1 that contains only the columns for Species, Sepal.Length, and Sepal.Width. How many observations and variables are in the dataset?

```{r q3}
# subset iris1 (question 2) for specific columns/variables
iris2 <- select(iris1,Species,Sepal.Length,Sepal.Width) 
# peek at new data frame
glimpse(iris2)
head(iris2)

```

There are 56 observations (rows) and 3 variables in the dataset.

### Question 4: Create an iris3 data frame from iris2 that orders the observations from largest to smallest sepal length. Show the first 6 rows of this dataset.

```{r q4}
iris3 <- arrange(iris2, by=desc(Sepal.Length)) #desc() changes order

#Show the first 6 rows of this dataset.
head(iris3) # head() defaults to the first 6 rows
```

### Question 5: Create an iris4 data frame from iris3 that creates a column with a sepal area (length * width) value for each observation. How many observations and variables are in the dataset?

```{r q5}
# create new variable called Sepal.Area (Sepal.Length*Sepal.Width) from iris3 data
iris4 <- mutate(iris3,Sepal.Area = Sepal.Length*Sepal.Width)
glimpse(iris4)
head(iris4)

```

There are 56 observations (rows) and 4 variables in the dataset

### Question 6: Create iris5 that calculates the average sepal length, the average sepal width, and the sample size of the entire iris4 data frame and print iris5.

```{r q6}
iris5 <- summarize(iris4, mean_length = mean(Sepal.Length), mean_width = mean(Sepal.Width), sample_size = n()) 

print(iris5)
```

### Question 7: 

Finally, create iris6 that calculates the average sepal length, the average sepal width, and the sample size for each species of in the iris4 data frame and print iris6.

```{r q7, message=FALSE}
iris6 <- group_by(iris4,Species) # by species

iris6 <- summarize(iris6, mean_length = mean(Sepal.Length), mean_width = mean(Sepal.Width), sample_size = n()) 

print(iris6)

```

### Question 8: 

In these exercises, you have successively modified different versions of the data frame iris1 iris1 iris3 iris4 iris5 iris6. At each stage, the output data frame from one operation serves as the input for the next.

A more efficient way to do this is to use the pipe operator %>% from the tidyr package. See if you can rework all of your previous statements into an extended piping operation that uses iris as the input and generates iris6 as the output.

```{r q8, message=FALSE}
iris %>%
  filter(Species %in% c("virginica","versicolor"),Sepal.Length > 6, Sepal.Width > 2.5) %>%
  select(Species,Sepal.Length,Sepal.Width) %>%
  arrange(by=desc(Sepal.Length)) %>%
  group_by(Species) %>%
  summarize(mean_length=mean(Sepal.Length),mean_width =mean(Sepal.Width), sample_size = n()) 

```

